Concurrent Programming in Harmony

### 2021-02-01 book/paper.tex
Pass over Chapter 11
Made pass over Ch 12

Locks are useful when accessing a shared data structure.  By preventing
more than one thread from accessing the data structure at the same
time, conflicting accesses are avoided.  However, not all concurrent
accesses conflict, and opportunities for concurrency may be lost,
hurting performance.  One important case is when multiple threads
are simply reading the data structure.
In many applications, reads are the majority of all accesses.
Allowing reads to proceed concurrently can significantly improve performance.

@@

\index{ring buffer}%
A \emph{bounded buffer} (aka \emph{ring buffer}) is
a queue with the usual \texttt{put/get} interface,
but implemented using a circular buffer
\index{circular buffer}%
of a certain length and two pointers:
the \emph{tail} points where new items are enqueued and the \emph{head}
points where items are dequeued.
If the buffer is full, the enqueuer must wait; if the buffer is empty, the
dequeuer must wait.
This problem is known as the ``Producer/Consumer Problem'' and was
proposed by Dijkstra~\cite{EWD329}.

The producer/consumer pattern is common.  Threads may be arranged
in \emph{pipelines},
\index{pipeline}%
where each upstream thread is a producer and each downstream
thread is a consumer.
Or threads may be arranged in a manager/worker pattern, with a manager
producing jobs and workers consuming and executing them in parallel.
Or, in the client/server model,
\index{client/server model}%
some thread may act as a \emph{server} that clients can send requests to
and receive responses from.  In that case, there is a bounded buffer
for each client/server pair. Clients produce requests and
consume responses, while the server consumes requests and produces responses.
%
Multiple producers and multiple consumers may all share
the same bounded buffer.

@@

Note that the following invariants hold:
\begin{itemize}
\item At any time, at most one gate is open;
\item If some gate is open, then no thread is in the critical section;
\item If some thread is in the critical section, all gates are closed;
\item At any time, at most one thread is in the critical section.
\end{itemize}

@@

We will illustrate the technique using the reader/writer problem.
\autoref{fig:RWsplitsema} shows code.
The first step is to enumerate all waiting conditions.
In the case of the reader/writer
problem, there are two: a thread that wants to read may have to wait for a
writer to leave the critical section, while a thread that wants to write may
have to wait until all readers have left the critical section or until a writer has left.
%
The state of a reader/writer lock thus consists of the following:
\begin{itemize}
\item \textit{nreaders}: the number of readers in the critical section;
\item \textit{nwriters}: the number of writers in the critical section;
\item \textit{mutex}: the ``main gate'' binary semaphore;
\item \textit{r\_gate}: the ``waiting gate'' used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
\item \textit{w\_gate}: the ``waiting gate'' used by writers, similar to the readers' gate.
\end{itemize}

@@

Let us examine \texttt{read\_acquire} more carefully.
First the method acquires the \textit{mutex}.
Then, in the case that the thread
finds that there is a writer in the critical section
($\mathit{nwriters > 0}$), it increments the counter associated with
the readers' gate, leaves the critical section (\texttt{release\_one}),
and then tries to acquire the semaphore associated with the gate.
This will cause it to block until some thread opens that gate.

Now consider the case where there is a writer in the critical
section and there are two readers waiting.  Let us see what happens when
the writer calls \texttt{write\_release}:
\begin{enumerate}
\item After acquiring \textit{mutex}, the writer decrements
\textit{nwriters}, which must be 1 at this time, and thus becomes 0.
\item It then calls \texttt{release\_one()}.
\texttt{release\_one()} finds that there are no writers in the critical section
and there are two readers waiting.  It therefore releases not
\textit{mutex} but the readers' gate's binary semaphore.
\item One of the waiting reader can now re-enter the critical section.
When it does, the reader decrements the gate's counter (from 2 to 1),
and increments \textit{nreaders} (from~0 to~1).
The reader finally calls \texttt{release\_one()}.
\item Again, \texttt{release\_one()} finds that there are no writers and
that there are readers waiting, so again it releases the readers' semaphore.
\item The second reader can now enter the critical section.
It decrements the gate's count from 1 to 0 and increments \textit{nreaders}
from~1 to~2.
\item Finally, the second reader  calls \texttt{release\_one()}.
This time \texttt{release\_one()} does not find any threads waiting,
and so it releases \textit{mutex}.
There are now two threads that are holding the reader/writer lock.
\end{enumerate}

\section*{Exercises}
\begin{problems}
% \item Assume that you have only binary semaphores.
% Implement counting semaphores using the split binary semaphores technique.
\item Several of the instantiations of \texttt{release\_one()} in
\autoref{fig:RWsplitsema} can be replaced by simply releasing \textit{mutex}.
Which ones?
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:bbsbs} Implement a solution to the producer/consumer problem
using split binary semaphores.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
(Bound locks are useful for situations where too many threads working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than \texttt{M} threads can acquire the
bound lock.
\item Write a test program for bound locks
that checks that up to \texttt{M} threads
can acquire the bound lock at the same time.
% \item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
% should block if the stack is full.  Method \texttt{pop()} should block
% until the stack is non-empty and then return an item.
% Also implement test programs for your solution.
\item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
\autoref{fig:gpu}.
There are \texttt{N} GPUs identified by the numbers
1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
of an available GPU, blocking if there is currently no available GPU.
Method \texttt{gpuRelease}(\textit{gpu}) releases the given GPU.  It never needs
to block.
\item With reader/writer locks,
concurrency can be improved if a thread \emph{downgrades} its write lock
to a read lock when its done writing but not done reading.  Add
a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
(Similarly, you may want to try to implement an \emph{upgrade} of a
read lock to a write lock.)
% How easy or hard would it be to do the same for the code in
% \autoref{fig:rw2lock}?
% Note, you cannot simply release the
% write lock and then obtain the read lock, because the thread needs
% to remain in the critical section.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.
Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
\texttt{nb\_enter()} that a car must invoke before going northbound on
the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most $n$ cars are allowed on the bridge.  Add $n$
as an argument to \texttt{OLBlock}.

>>>>

Locks are useful when accessing a shared data structure.  By preventing
more than one thread from accessing the data structure at the same
time, conflicting accesses are avoided.  However, not all concurrent
accesses conflict, and opportunities for concurrency may be lost,
hurting performance.  One important case is when multiple threads
are simply reading the data structure.
In many applications, reads are the majority of all accesses,
and read operations do not conflict with one another.
Allowing reads to proceed concurrently can significantly improve performance.

@@

\index{ring buffer}%
A \emph{bounded buffer} (aka \emph{ring buffer}) is
a queue with the usual \texttt{put/get} interface,
but implemented using a circular buffer
\index{circular buffer}%
of a certain length and two pointers:
the \emph{tail} points where new items are enqueued and the \emph{head}
points where items are dequeued.
If the buffer is full, the enqueuer must wait; if the buffer is empty, the
dequeuer must wait.
This problem is known as the ``Producer/Consumer Problem'' and was
proposed by Dijkstra~\cite{EWD329}.
Multiple producers and multiple consumers may all share
the same bounded buffer.

The producer/consumer pattern is common.  Threads may be arranged
in \emph{pipelines},
\index{pipeline}%
where each upstream thread is a producer and each downstream
thread is a consumer.
Or threads may be arranged in a manager/worker pattern, with a manager
producing jobs and workers consuming and executing them in parallel.
Or, in the client/server model,
\index{client/server model}%
some thread may act as a \emph{server} that clients can send requests to
and receive responses from.  In that case, there is a bounded buffer
for each client/server pair. Clients produce requests and
consume responses, while the server consumes requests and produces responses.

@@

Note that the following invariants hold:
\begin{itemize}
\item At any time, at most one gate is open;
\item If some gate is open, then no thread is in the critical section.  Equivalently, if some thread is in the critical section, all gates are closed;
\item At any time, at most one thread is in the critical section.
\end{itemize}

@@

We will illustrate the technique using the reader/writer problem.
\autoref{fig:RWsplitsema} shows code.
The first step is to enumerate all waiting conditions.
In the case of the reader/writer
problem, there are two: a thread that wants to read may have to wait for a
writer to leave the critical section, while a thread that wants to write may
have to wait until all readers have left the critical section or until a writer has left.
%
The state of a reader/writer lock thus consists of the following:
\begin{itemize}
\item \textit{nreaders}: the number of readers in the critical section;
\item \textit{nwriters}: the number of writers in the critical section;
\item \textit{mutex}: the main gate binary semaphore;
\item \textit{r\_gate}: the waiting gate used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
\item \textit{w\_gate}: the waiting gate used by writers, similar to the readers' gate.
\end{itemize}

@@

Let us examine \texttt{read\_acquire} more carefully.
First the method acquires the \textit{mutex}.
Then, in the case that the thread
finds that there is a writer in the critical section
($\mathit{nwriters > 0}$), it increments the counter associated with
the readers' gate, leaves the critical section (\texttt{release\_one}),
and then tries to acquire the semaphore associated with the waiting gate.
This will cause it to block until some thread opens that gate.

Now consider the case where there is a writer in the critical
section and there are two readers waiting.  Let us see what happens when
the writer calls \texttt{write\_release}:
\begin{enumerate}
\item After acquiring \textit{mutex}, the writer decrements
\textit{nwriters}, which must be 1 at this time, and thus becomes 0.
\item It then calls \texttt{release\_one()}.
\texttt{release\_one()} finds that there are no writers in the critical section
and there are two readers waiting.  It therefore releases not
\textit{mutex} but the readers' gate's binary semaphore.
\item One of the waiting readers can now re-enter the critical section.
When it does, the reader decrements the gate's counter (from 2 to 1)
and increments \textit{nreaders} (from~0 to~1).
The reader finally calls \texttt{release\_one()}.
\item Again, \texttt{release\_one()} finds that there are no writers and
that there are readers waiting, so again it releases the readers' semaphore.
\item The second reader can now enter the critical section.
It decrements the gate's count from 1 to 0 and increments \textit{nreaders}
from~1 to~2.
\item Finally, the second reader  calls \texttt{release\_one()}.
This time \texttt{release\_one()} does not find any threads waiting,
and so it releases \textit{mutex}.
There are now two threads that are holding the reader/writer lock.
\end{enumerate}

\section*{Exercises}
\begin{problems}
% \item Assume that you have only binary semaphores.
% Implement counting semaphores using the split binary semaphores technique.
\item Several of the instantiations of \texttt{release\_one()} in
\autoref{fig:RWsplitsema} can be replaced by simply releasing \textit{mutex}.
Which ones?
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:bbsbs} Implement a solution to the producer/consumer problem
using split binary semaphores.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
(Bound locks are useful for situations where too many threads working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than \texttt{M} threads can acquire the
bound lock.
\item Write a test program for bound locks
that checks that up to \texttt{M} threads
can acquire the bound lock at the same time.
% \item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
% should block if the stack is full.  Method \texttt{pop()} should block
% until the stack is non-empty and then return an item.
% Also implement test programs for your solution.
\item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
\autoref{fig:gpu}.
There are \texttt{N} GPUs identified by the numbers
1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
of an available GPU, blocking if there is currently no available GPU.
Method \texttt{gpuRelease}(\textit{gpu}) releases the given GPU.  It never needs
to block.
\item With reader/writer locks,
concurrency can be improved if a thread \emph{downgrades} its write lock
to a read lock when its done writing but not done reading.  Add
a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
(Similarly, you may want to try to implement an \emph{upgrade} of a
read lock to a write lock.  Why is this problematic?)
% How easy or hard would it be to do the same for the code in
% \autoref{fig:rw2lock}?
% Note, you cannot simply release the
% write lock and then obtain the read lock, because the thread needs
% to remain in the critical section.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.
Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
\texttt{nb\_enter()} that a car must invoke before going northbound on
the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most $n$ cars are allowed on the bridge.  Add $n$
as an argument to \texttt{OLBlock}.
